Transcriptomic responses to diet quality and viral infection in Apis mellifera

### 2018-11-14 Manuscript/Text/manuscript.tex
Toth additional revisions

Managed honey bees have undergone unusually large declines in the United States and parts of Europe over the past decade \cite{ccd1, ccd2, ccd3}, with annual mortality rates exceeding what beekeepers consider sustainable \cite{ccd5, ccd6}. More than 70 percent of major global food crops (including fruits, vegetables, and nuts) at least benefit from pollination, and yearly insect pollination services are valued worldwide at \$175 billion \cite{ccd7}. As honey bees are largely considered to be the leading pollinator of numerous crops, their marked loss has considerable implications for agricultural sustainability \cite{ccd4}.

@@

There are only a small number of published experiments examining gene expression patterns related to diet effects \cite{alaux2} and virus infection effects \cite{galbraith} in honey bees, but there have been several such studies in model organisms. Model insect studies can inform studies of honey bee transcriptomic responses, using functional inference of as-of-yet uncharacterized honey bee genes based on orthology to Drosophila and other model organisms. Previous Drosophila studies that examined various diet effects have found gene expression changes related to immunity, metabolism, cell cycle activity, DNA binding, transcription, and insulin signaling \cite{sugarFat, sugarProtein, jHormone, alaux2}. While similar transcriptomic studies have been limited in honey bees, one study found that pollen nutrition upregulates genes involved in macromolecule metabolism, longevity, and the insulin/TOR pathway required for physiological development \cite{alaux2}. Numerous studies on the transcriptomic effects of virus infection in model insect organisms have shown that RNA silencing, transcriptional pausing, Toll pathways, IMD pathways, JAK/STAT pathways, and Toll-7 autophagy pathways play substantial roles in virus-host systems \cite{cherryReview, swevers}. Virus-bee systems have also revealed key factors of these antiviral conserved defense pathways \cite{McMenamin}.

As far as we know, there are few to no studies investigating honey bee gene expression patterns specifically related to monofloral diets, and few studies investigating honey bee gene expression patterns related to the combined effects of diet in any broad sense and viral inoculation in any broad sense \cite{interactingDV}. In this study, we examine how monofloral diets and viral inoculation influence gene expression patterns in honey bees by focusing on four treatment groups (low quality diet without IAPV exposure, high quality diet without IAPV exposure, low quality diet with IAPV exposure, and high quality diet with IAPV exposure). For our diet factor, we examined two monofloral pollen diets, rockrose (\textit{Cistus} sp.) and chestnut (\textit{Castanea} sp.). Rockrose pollen is generally considered less nutritious than chestnut pollen because it contains smaller amounts of protein, amino acids, antioxidants, calcium, and iron \cite{adamInt, DiPasquale}. We conduct RNA-sequencing analysis on a randomly selected subset of the honey bees we used in our previous study (as is further described in our methods section). We then examine pairwise combinations of treatment groups, the main effect of monofloral diet, the main effect of IAPV exposure, and the combined effect of the two factors on gene expression patterns.

We also compare the main effect of IAPV exposure in our dataset to that obtained in a previous study conducted by Galbraith and colleagues \cite{galbraith}. While our study examines honey bees derived from naturally-mated queens, the Galbraith study examined honey bees derived from single-drone inseminated queens. As a consequence, the honey bees in our study will be on average 25\% genetically identical, whereas honey bees from the Galbraith study will be on average 75\% genetically identical \cite{sisters}. We note that the difference between these studies may be even greater than this as we used honey bees from 15 different colonies, i.e. from 15 different, naturally-mated queens. We should therefore expect that the Galbraith study may generate data with higher signal:to:noise ratios than our data due to lower genetic variation between its replicates. At the same time, our honey bees will be more likely to display the health benefits gained from increased genotypic variance within colonies, including decreased parasitic load \cite{multParasite}, increased tolerance to environmental changes \cite{divHyp2}, and increased colony performance \cite{geneticDiverse, geneticDiverse2}. Given that honey bees are naturally very polyandrous \cite{patriline}, our naturally-mated honey bees may also reflect more realistic environmental and genetic conditions. Taken together, each study provides a different point of value: Our study likely presents less artificial data while the Galbraith data likely presents less messy data. We therefore wish to explore how the gene expression effects of IAPV inoculation compare between these two studies using different experimental designs. To achieve this objective, we use visualization techniques to assess the signal:to:noise ratio between these two datasets, and differential gene expression (DEG) analyses to determine any significantly overlapping genes of interest between these two datasets. As RNA-sequencing data can be biased \cite{biased1, biased2, biased3}, this comparison allowed us to characterize how repeatable and robust our RNA-sequencing results were in comparison to previous studies. It also allowed us to shine light on how experimental designs that control genetic variability to different extents might affect the resulting gene expression data in honey bees. We suggest that in-depth data visualization approaches can be useful for cross-study comparisons and validation of noisy RNA-sequencing data in the future.

@@

We wished to explore the signal:to:noise ratio between the Galbraith dataset and our dataset. Note that the Galbraith dataset contained three samples for each virus level, while our dataset contained twelve samples for each virus level. Basic PCA plots were constructed with the DESeq2 analysis pipeline and showed that the Galbraith dataset may separate the infected and uninfected honey bees better than our dataset (Additional file 2). We also noted that the first replicate of both treatment groups in the Galbraith data did not cluster as cleanly in the PCA plots. However, through this automatically-generated plot, we can only visualize information at the sample level. Wanting to learn more about the data at the gene level, we continued with new visualization techniques that are available online (https://lrutter.github.io/bigPint) and are in preparation for publication.

We used parallel coordinate lines superimposed onto side-by-side boxplots to visualize the DEGs associated with virus infection in the two studies. The background side-by-side boxplot represents the distribution of \textit{all} genes in the data, and each parallel coordinate line represents one DEG. In a parallel coordinate line, connections between samples with positive correlations should be flat, while connections between samples with negative correlations should be crossed. We expect DEGs to show more variability between treatments than between replicates. This means the parallel coordinate lines should be flat between replicates but crossed between treatments. However, overplotting problems would obscure our visualization if we were to plot all DEGs onto the same side-by-side boxplot. As a result, we used hierarchical clustering techniques to separate DEGs into common patterns as is described in the methods section.

We see that the 1,019 DEGs from the Galbraith dataset form relatively clean-looking visual displays, with consistent replicates and differences between treatments (Figure 3). We do see that the first replicate of the virus group (V.1) appears somewhat inconsistent with the other virus replicates in Cluster 1, confirming that the trend we saw in the PCA plot carried through into the DEG results. Cluster 4 reveals somewhat inconsistent replicates in the virus group, although most virus standardized read counts (group V) remain consistently larger than most control standardized read counts (group N). In contrast, we see that the 43 virus-related DEGs from our dataset do not look as clean in their visual displays (Figure 4). The replicates appear somewhat inconsistent in their estimated expression levels and there is not always such a large (or even consistent) difference between treatment groups. We see a similar finding when we also examine a larger subset of 1,914 diet-related DEGs from our study (Additional file 3).

We next used repLIcate TREatment (``litre'') plots, which we recently developed and published in our bigPint software package. Litre plots allow users to visualize one DEG onto the Cartesian coordinates of one scatterplot matrix. In the litre plot, each gene in the data is plotted once for every combination of replicates between treatment groups. For example, there are nine ways to pair a replicate from one treatment group with a replicate from the other treatment group in the Galbraith dataset (N.1 and V.1, N.1 and V.2, N.1 and V.3, N.2 and V.1, N.2. and V.2, N.2 and V.3, N.3 and V.1, N.3 and V.2, and N.3 and V.3). Hence, each gene in the Galbraith dataset is plotted as nine points in the litre plot. With 11,825 genes in the Galbraith data, 106,425 points would need to be plotted. Our dataset is even more dramatic: There are 144 ways to pair a replicate from one treatment group with a replicate from the other treatment group, and with 15,314 genes in our data, we would need to plot 2,205,216 points. For either dataset, plotting all these points would reduce the speed of interactive functionality and cause overplotting problems. As a result, we use hexagon bins to summarize this massive information. Once the background of hexagons has been drawn to reveal the distribution of all between-treatment sample pair combinations for \textit{all} genes, the user can superimpose all between-treatment sample pair combinations for one gene of interest.

Additional file 4 shows nine example litre plots for our dataset. The hexagon background is the same for all nine litre plots because it simply shows the distribution of all between-treatment sample pair combinations for \textit{all} genes in our dataset. In each litre plot, there are 144 magenta points superimposed that show all between-treatment sample pair combinations for one DEG of interest. Additional file 5 and 6 similarly each show nine example litre plots for the Galbraith dataset. We examined individual DEGs from the first cluster (Additional file 5) and second cluster (Additional file 6) of the Galbraith data because the first cluster had previously shown less consistency in the first replicate of the treatment group (Figure 3). Notice that, as previously explained, we now show each DEG as nine points for the Galbraith dataset. We see that indeed the virus DEGs from our data (Additional file 4) show less consistent replications and less differences between the treatment groups compared to the virus DEGs from the Galbraith data (Additional files 5 and 6). We also observe that, in the Galbraith dataset, the DEG points in the first cluster show less tight cluster patterns than the DEG points in the second cluster (Additional files 5 and 6), an observation we saw previously in the parallel coordinate plots (Figure 3).

>>>>

Managed honey bees have undergone health declines in the United States and parts of Europe over the past decade \cite{ccd1, ccd2, ccd3}, with annual mortality rates exceeding what beekeepers consider sustainable \cite{ccd5, ccd6}. More than 70 percent of major global food crops (including fruits, vegetables, and nuts) at least benefit from pollination, and yearly insect pollination services are valued worldwide at \$175 billion \cite{ccd7}. As honey bees are largely considered to be the leading pollinator of numerous crops, their marked loss has considerable implications for agricultural sustainability \cite{ccd4}.

@@

There are only a small number of published experiments examining gene expression patterns related to diet effects \cite{alaux2} and virus infection effects \cite{galbraith} in honey bees, but there have been several such studies in model organisms. Model insect studies can inform studies of honey bee transcriptomic responses, using functional inference of as-of-yet uncharacterized honey bee genes based on orthology to Drosophila and other model organisms. Previous Drosophila studies that examined various diet effects have found gene expression changes related to immunity, metabolism, cell cycle activity, DNA binding, transcription, and insulin signaling \cite{sugarFat, sugarProtein, jHormone, alaux2}. While similar transcriptomic studies have been limited in honey bees, one study found that pollen nutrition upregulates genes involved in macromolecule metabolism, longevity, and the insulin/TOR pathway required for physiological development \cite{alaux2}. Numerous studies on the transcriptomic effects of virus infection in model insect organisms have shown that RNA silencing, transcriptional pausing, Toll pathways, IMD pathways, JAK/STAT pathways, and Toll-7 autophagy pathways play substantial roles in virus-host systems \cite{cherryReview, swevers}. Studies of virus-bee systems have revealed some of the antiviral defense pathways known in model organisms are conserved and also related to bee antiviral immune responses \cite{McMenamin}.

To our knowledge, there are few to no studies investigating honey bee gene expression patterns specifically related to monofloral diets, and few studies investigating honey bee gene expression patterns related to the combined effects of diet in any broad sense and viral inoculation in any broad sense \cite{interactingDV}. In this study, we examine how monofloral diets and viral inoculation influence gene expression patterns in honey bees by focusing on four treatment groups (low quality diet without IAPV exposure, high quality diet without IAPV exposure, low quality diet with IAPV exposure, and high quality diet with IAPV exposure). For our diet factor, we examined two monofloral pollen diets, rockrose (\textit{Cistus} sp.) and chestnut (\textit{Castanea} sp.). Rockrose pollen is generally considered less nutritious than chestnut pollen because it contains smaller amounts of protein, amino acids, antioxidants, calcium, and iron \cite{adamInt, DiPasquale}. We conduct RNA-sequencing analysis on a randomly selected subset of the honey bees we used in our previous study (as is further described in our methods section). We then examine pairwise combinations of treatment groups, the main effect of monofloral diet, the main effect of IAPV exposure, and the combined effect of the two factors on gene expression patterns.

Because RNA-seq data can be noisy and subject to high levels of inter-experiment variation, we further sought to validate our transcriptomic data via comparison to a previous RNA-seq study on honey bee responses to viral infection. To do this, we compare the main effect of IAPV exposure in our dataset to that obtained in a previous study conducted by Galbraith and colleagues \cite{galbraith}. While our study examines honey bees derived from naturally-mated queens, the Galbraith study examined honey bees derived from single-drone inseminated queens. As a consequence, the honey bees in our study will be on average 25\% genetically identical, whereas honey bees from the Galbraith study will be on average 75\% genetically identical \cite{sisters}. We note that the difference between these studies may be even greater than this as we used honey bees from 15 different colonies, i.e. from 15 different, naturally-mated queens. We should therefore expect that the Galbraith study may generate data with higher signal:to:noise ratios than our data due to lower genetic variation between its replicates. At the same time, our honey bees will be more likely to display the health benefits gained from increased genotypic variance within colonies, including decreased parasitic load \cite{multParasite}, increased tolerance to environmental changes \cite{divHyp2}, and increased colony performance \cite{geneticDiverse, geneticDiverse2}. Given that honey bees are naturally very polyandrous \cite{patriline}, our naturally-mated honey bees may also reflect more realistic environmental and genetic conditions. Taken together, each study provides a different point of value: Our study likely presents less artificial data while the Galbraith data likely presents less messy data. To achieve this comparison, we use visualization techniques to assess the signal:to:noise ratio between these two datasets, and differential gene expression (DEG) analyses to determine any significantly overlapping genes of interest between these two datasets. As RNA-sequencing data can be biased \cite{biased1, biased2, biased3}, this comparison allowed us to characterize how repeatable and robust our RNA-sequencing results were in comparison to previous studies. It also allowed us to shine light on how experimental designs that control genetic variability to different extents might affect the resulting gene expression data in honey bees. We suggest that in-depth data visualization approaches can be useful for cross-study comparisons and validation of noisy RNA-sequencing data in the future.

@@

We wished to explore the signal:to:noise ratio between the Galbraith dataset and our dataset. Note that the Galbraith dataset contained three samples for each virus level, while our dataset contained twelve samples for each virus level. Basic PCA plots were constructed with the DESeq2 analysis pipeline and showed that the Galbraith dataset may separate the infected and uninfected honey bees better than our dataset (Additional file 2). Wanting to learn more about the data at the gene level, we continued with new visualization techniques that are available online (https://lrutter.github.io/bigPint). To understand how to interpret the visualizations used in this section, please read (https://lrutter.github.io/bigPint/articles/plotIntro.html).

We used parallel coordinate lines superimposed onto side-by-side boxplots to visualize the DEGs associated with virus infection in the two studies. The background side-by-side boxplot represents the distribution of \textit{all} genes in the data, and each parallel coordinate line represents one DEG. In a parallel coordinate line, connections between samples with positive correlations should be flat, while connections between samples with negative correlations should be crossed. We expect DEGs to show more variability between treatments than between replicates. This means the parallel coordinate lines should be flat between replicates but crossed between treatments. However, overplotting problems would obscure our visualization if we were to plot all DEGs onto the same side-by-side boxplot. Therefore, we graphed clustered subsets of the DEGs (based on hierarchical clustering).

The 1,019 DEGs from the Galbraith dataset form relatively clean-looking visual displays, with consistent replicates and differences between treatments (Figure 3). In contrast, we see that the 43 virus-related DEGs from our dataset do not look as clean in their visual displays (Figure 4). The replicates appear somewhat inconsistent in their estimated expression levels and there is not always such a large (or even consistent) difference between treatment groups. We see a similar finding when we also examine a larger subset of 1,914 diet-related DEGs from our study (Additional file 3).

We next used repLIcate TREatment (``litre'') plots, which we recently developed for our bigPint software package. Litre plots allow users to visualize one DEG onto the Cartesian coordinates of one scatterplot matrix. In the litre plot, each gene in the data is plotted once for every combination of replicates between treatment groups. We use hexagon bins to summarize this massive information. Once the background of hexagons has been drawn to reveal the distribution of all between-treatment sample pair combinations for \textit{all} genes, the user can superimpose all between-treatment sample pair combinations for one gene of interest.

Additional file 4 shows nine example litre plots for our dataset; each litre plot shows the 144 between-treatment sample pair combinations for one DEG of interest. Additional file 5 and 6 similarly each show nine example litre plots for the Galbraith dataset; each litre plot shows the nine between-treatment sample pair combinations for one DEG of interest. We see that indeed the virus DEGs from our data (Additional file 4) show less consistent replications and less differences between the treatment groups compared to the virus DEGs from the Galbraith data (Additional files 5 and 6). We also observe that, in the Galbraith dataset, the DEG points in the first cluster show less tight cluster patterns than the DEG points in the second cluster (Additional files 5 and 6), an observation we saw previously in the parallel coordinate plots (Figure 3).
